Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2017-12-08 paper.tex
updated main graph, comments to text a

\def\RA{A$_{1}$}
\def\RB{A$_{2}$}

@@

Two motion planners were compared: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
The \RA\ method was run with the following parameters:
$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
Both methods utilize the same data structures and low-level routines (e.g., collision-detection) to enable fair comparison.
%were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners as it resulted in the highest success rates.

@@

In the following experiments, we measured the performance of the methods using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA.~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}

@@

The aim of the first experiment is to measure the performance of both planners in the simplest scenario, i.e., finding a path for one sphere heading towards the active site.
The radius of the sphere is set to $1.7$~\AA.

The computed traversability rates are shown in Tab.~\ref{tab::one}.
The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with the bottleneck larger than the tested radius.
The row 'Planning' shows the traversable rate computed by the planners.
This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been implemented correctly.
Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
increase the planning time.
For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.

@@

Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane), and
the task was to compute the trajectories for these ligands in a given molecular dynamics simulation.
Models of the ligands in Mol2 in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand} containing the topology, rotatable bonds, atom types and partial charges. 
Rotamer library was created by rotation of all rotatable bonds with 30$^\circ$ step. 
The energy cut-off for individual conformations was set to 0.075 %(-rot_ensemble_ecutoff 0.075) 
and Rosetta database 3.2.1 was used during the conformers generation.

@@

The results are summarized in Tab.~\ref{tab::main}.
The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.

\begin{table}[bt]
\centering
\caption{\label{tab::main}
    \small
    Traversability of the tunnels for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\footnotesize
\renewcommand{\arraystretch}{0.7}
\input graph.tex
}
\end{table}

The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (8~\%), m037t (20~\%) and m038t (16~\%).

In conclusion, the traversability of the DCP ligand to active site was approved by the MD simulation~\cite{marques2017catalytic}.
The observed trajectory of the DCP ligand in the MD simulation followed the main tunnel, as expected.
Based on the traversability rates with $\smin=0.8$, we expect that at least ligands m037t and m038t can also reach the active site, as
they have the same (or even larger) traversability rate than the DCP.
However, this hypothesis has not been approved yet by testing it on MD simulations.

@@

>>>>

\def\RA{A$_{1}$}
\def\RB{A$_{2}$}
\def\RC{A$_{1}^{*}$}
\def\RD{A$_1^P$}

@@

Two motion planners were compared: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
The \RA\ method was run with the following parameters:
$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
Both methods utilize the same data structures and low-level routines (e.g., collision-detection) to enable fair comparison.
%were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners as it resulted in the highest success rates.
The \RA\ method was also tested without the atom-based metric.
In this case (denoted \RC), the distance in the expansion step is measured using 6D Euclidean metric.

@@

In the following experiments, we measured the performance of the methods using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA. 
%~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}

@@

The performance of \RA\ and \RB\ planners is first measured by finding trajecotries for only one atom of VDW radii $1.7$~\AA.
The \RC\ variant is not tested here as it behaves same as \RA\ in the case of one-atom ligand.
The computed traversability rates are shown in Tab.~\ref{tab::one}.
The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with the bottleneck larger than the tested radius.
The row 'Planning' shows the traversable rate computed by the planners.
This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been implemented correctly.
Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
increase the planning time.
For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.

@@

Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane).
Models of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand} containing the topology, rotatable bonds, atom types and partial charges. 
Rotamer library was created by rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
%and Rosetta database 3.2.1 was used during the conformers generation.

@@

\begin{table}[bt]
\centering
\caption{\label{tab::main}
    \small
    Traversability rate [\%] over 100 frames for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
    Each frame of protein dynamics consists of 4650 atoms.
}
\small
\renewcommand{\tabcolsep}{2pt}
{
\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphNew.tex
}
\end{table}

%{
%\footnotesize
%\begin{table}
%\caption{New}
%\renewcommand{\tabcolsep}{2pt}
%\scriptsize
%\input g2.tex
%\end{table}
%}

The results are summarized in Tab.~\ref{tab::main}.
The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the ligand DCP, trajectories were detected in 10~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
%For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (18~\%), m037t (45~\%) and m038t (30~\%).
%The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
Based on the number of success rate, also other ligands (e.g., m037t with trav. rate 45~\% at $\smin$=0.8) seem to be promissing.
However, their binding was not approved yet by MD dedicated simulations.

%The observed trajectory of the DCP ligand in the MD simulation followed the main tunnel, as expected.
%Based on the traversability rates with $\smin=0.8$, we expect that at least ligands m037t and m038t can also reach the active site, as
%they have the same (or even larger) traversability rate than the DCP.
%However, this hypothesis has not been approved yet by testing it on MD simulations.

@@

Beside scaling-down the ligand, also the atoms of the protein can be scaled-down, which is used e.g. in~\cite{cortes2005path} which
scales  atoms of both the ligand and the protein to 80~\% of their size.
We tested also this case (column \RD\ in Tab.~\ref{tab::main}).
The results are same as for ligand scale $\smin=0.6$ (except DCP and m040 which can be results of the randomization).
